Forecasting Influenza Hospitalizations Using a Bayesian Hierarchical Nonlinear Model with Discrepancy

### 2025-06-30 nonlinear_forecast.tex
reformat sections
fix full fit

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only one address is permitted per author. %%
%% Only division, organization and e-mail is %%
%% included in the address.                  %%
%% Additional information can be included in %%
%% the Acknowledgments section if necessary. %%
%% ORCID can be inserted by command:         %%
%% \orcid{0000-0000-0000-0000}               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\snm{Spencer Wadsworth}~\ead[label=e1]{sgw96@iastate.edu}}
% \author[A]{\fnms{Second}~\snm{Author}\ead[label=e2]{second@somewhere.com}\orcid{0000-0000-0000-0000}}
\and
\author[A]{~\snm{Jarad Niemi}\ead[label=e2]{niemi@iastate.edu}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Department of Statistics,
Iowa State University, Ames, IA 50011}

@@

\begin{abstract}
The annual influenza outbreak leads to significant public health and economic burdens making it desirable to have prompt and accurate probabilistic forecasts of the disease spread. The United States Centers for Disease Control and Prevention (CDC) hosts annually a national flu forecasting competition which has led to the development of a variety of flu forecast modeling methods. For the first several years of the competition, the target to be forecast was weekly percentage of patients with an influenza-like illness (ILI), but in 2021 the target was changed to weekly hospitalization counts. Reliable state and national hospitalization data has only been available since 2021, but for ILI the data has been available since 2010 and has been successfully forecast for several seasons.
In this manuscript, we introduce a two component modeling framework for forecasting weekly hospitalizations utilizing both hospitalization data and ILI data. The first component is for modeling ILI data using a dynamic nonlinear Bayesian hierarchical model. The second component is for modeling hospitalizations as a function of ILI. For hospitalization forecasts, ILI is first forecasted and then hospitalizations are forecast with ILI forecasts used as a linear or quadratic predictor. In a simulation study, two ILI forecast models, including one similar to the winning model for two seasons of the CDC forecast competition from \cite{osthus2019dynamic} and a nonlinear Bayesian hierarchical model from \cite{ulloa2019} are compared. Also assessed is the usefulness of including a systematic model discrepancy term in the ILI model. Forecasts of state and national hospitalizations for the 2023-24 flu season are made, and different modeling decisions are compared. We found that including a discrepancy component in the ILI model tends to improve forecasts during certain weeks of the year. We also found that other modeling decisions such as the exact nonlinear function to use in the ILI model or the error distribution for hospitalization models may or may not be better than other decisions, depending on the season, location, or week of the forecast.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% `\ ' is used here because TeX ignores    %%
%% spaces after text commands.              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

Every year the seasonal influenza outbreak burdens the public health system by infecting millions, causing an influx of primary care visits and hospitalizations and leading to between 290,000 and 650,000 deaths worldwide \cite[]{whoflufact2023}. \cite{molinari2007annual} estimated the United States' annual economic burden from medical costs, loss of income, and deaths to be over \$87 billion. Accurate forecasting of infectious diseases can inform public decision making and ease the burden of an outbreak \cite[]{turtle2021accurate, lutz2019applying}.
There is a growing consensus that disease forecasts should be probabilistic in nature \cite[]{gneiting2014probabilistic, bracher2021evaluating}, and it has been shown that reporting forecast uncertainty along with predictions may lead to better decision making \cite[]{ramos2013probabilistic, joslyn2012uncertainty, winkler1971probabilistic}.

% \jarad{Test}
% \spencer{Test}

To better inform public decision making regarding the flu epidemic, in 2013 the United States Centers for Disease Control and Prevention (CDC) organized a national flu forecasting competition, also known as FluSight \cite[]{biggerstaff2016results, mathis2024evaluation,cdcfluforecasting2024}. Originally, over a dozen teams of researchers from academic and industry backgrounds participated in FluSight by contributing their own forecast models. Besides the 2020 season --or the flu season spanning the fall of 2020 and the winter of 2021-- FluSight has been operated annually and researchers outside the CDC have been invited to participate. Initially the target data for forecasts was influenza-like illness (ILI) data. ILI is the proportion of patients who meet a healthcare provider and who display flu like symptoms, and ILI data has been available at the state and national level since the 2010 flu season \cite[]{cdcfluview2023,cdc2024fluviewportal}. 
The collaborative ILI forecasting effort has led to a number of modeling developments in flu forecasting \cite[see references therein for more examples]{mcandrew2021adaptively, osthus2021multiscale, osthus2019dynamic, ulloa2019}, and in their paper's introduction, \cite{osthus2019dynamic} categorized the most commonly used flu forecasting models into four classes including mechanistic models based on differential equation compartmental models, agent based models based on population simulation, machine learning/regression models including data driven machine learning and statistical models, and data assimilation models which are constructed by assimilating mechanistic models into a probabilistic framework. An additional forecast model used in FluSight involves the combination of several forecasts into a single ensemble forecast, which has been shown to perform well relative to individual models \cite[]{mcandrew2021adaptively, ray2020ensemble, yamana2016superensemble}.

The administration of FluSight saw few changes during the first seven seasons, but the onset of the COVID-19 pandemic and subsequent developments for COVID-19 forecasting led to major modifications. As a result of the COVID-19 pandemic which began during the 2019 flu season, the typical flu outbreak behavior was altered between the 2019 and 2022 seasons \cite[]{mathis2024evaluation}. The COVID-19 pandemic led to the creation of the Health and Human Services (HHS) Patient Impact and Hospital Capacity Data System \cite[]{healthdata2024covidts} which contains COVID-19 and flu hospitalization data, and the COVID-19 Forecast Hub was founded. The COVID-19 Forecast Hub was based on FluSight but with certain major adjustments including how the forecast uncertainty is represented and the addition of the weekly publication of a multi-model ensemble forecast as the official forecast of the CDC \cite[]{bracher2021evaluating, Cramer2022-hub-dataset}. Using estimated quantiles for representing forecast uncertainty and creating a multi-model ensemble are both aspects of the COVID-19 Forecast Hub which were adopted by the flu forecast competition. Additionally the target of the flu forecasts changed from being ILI data to being HHS hospitalization data, which reports the number of hospitalizations due to a laboratory confirmed flu infection  \cite[]{mathis2024evaluation,healthdata2024covidts}. This is as a result of having COVID-19 cases in the population making ILI data, already only a proxy for flu behavior, more difficult to interpret.

The contribution of this manuscript is to introduce a two component framework for modeling HHS hospitalization forecasts where hospitalization data and years of ILI data are used to inform forecast models. The first modeling component is a model of ILI data and the second is a model of hospitalization data with ILI as a predictive covariate.
Herein we use ILI models similar to those in \cite{osthus2019dynamic} and \cite{ulloa2019} for ILI forecasting.
The model of \cite{osthus2019dynamic} is a combined data assimilation and statistical regression model which involves a compartmental model in a probabilistic framework. Their model also includes an additional component for capturing a systematic discrepancy between the deterministic part of the model and the actual data, an idea which was first introduced by \cite{kennedy2001bayesian}. The model in \cite{ulloa2019} is a Bayesian hierarchical regression model with an underlying function intended to capture the trajectory of the seasonal ILI data. Herein, we provide a framework under which discrepancy modeling may be used along with a general function modeling ILI data, and we show the effectiveness of including discrepancy modeling during certain periods of the flu season.

In line with the newer FluSight standard of forecasting hospitalizations, we model hospitalizations as a linear function of ILI. Thus forecasts produced herein target flu hospitalizations and are a mapping of ILI forecasts to hospitalizations. This allows for ILI data from many seasons to be exploited and for ILI forecasts to assist in forecasting hospitalizations, which has fewer seasons of data than ILI. Several modeling schemes and their forecasts for the 2023 flu hospitalization season are compared, and it is shown that the modeling decisions produce good forecast results for different states or times during the flu season.

In section \ref{sec:data} we review the ILI and hospitalization data provided by the CDC and targeted by FluSight. 
In section \ref{sec:functions} the modeling framework contributed by this manuscript is given. In the same section, functions similar to those used by \cite{osthus2019dynamic} and \cite{ulloa2019} are defined.
 These functions are the susceptible-infectious-recovered (SIR) compartmental model and the asymmetric Gaussian (ASG) function  respectively. Model fitting and implementation are described at the end of the section. Section \ref{sec:simulation2} is a simulation study where four ILI forecast models and their use in forecasting hospitalizations are compared. Commonly used proper scoring rules \cite[]{gneiting2007strictly}, which are also introduced and defined in section \ref{sec:simulation2}, are used for comparing the forecasts. Forecasting of the 2023 flu outbreak along with assessment and comparison under several selected models is performed in section \ref{sec:analysis}, with the analysis being done using the conventions of FluSight. 
 Finally, the manuscript is concluded in section \ref{sec:conclusion} with general observations and some discussion.

 \section{Flu outbreak data} \label{sec:data}
In this section we introduce and define ILI and hospitalization data and evaluate the data visually. ILI and hospitalization data have been the object of forecasting for FluSight with ILI being the target for the first seven seasons and hospitalizations being the target since the 2022 season. Both of these data were collected at the state, territorial, and national level and were reported at least weekly. Overall the data is reported for 53 locations including the 50 US states, the District of Columbia (DC), Puerto Rico (PR), and at the US national level. We will refer to forecast targets throughout this manuscript. A target is the specific horizon, 1, 2, 3, or 4-weeks ahead, for a specific location and week during the season.

\subsection{Influenza-like illness data}
The US Outpatient Influenza-like Illness Surveillance Network (ILINet) collects information on respiratory illness from outpatient visits to health care providers. Over 3,400 outpatient health care providers in all 50 US states, PR, DC, and the US Virgin Islands report each week the total number of outpatient visits along with the number of ILI cases. An ILI case is defined as a ``fever (temperature of $100^{\circ}$F[$37.8^{\circ}$C] or greater) and a cough and/or a sore throat." Prior to the 2021 season, the definition included ``without a known cause other than influenza" \cite[]{cdcfluview2023}. Because other illnesses such as COVID-19, RSV, and the common cold may induce similar respiratory symptoms, ILI may include patients infected with some disease other than influenza. To know whether or not a sick patient is infected with influenza would require a laboratory test.

In 2013, when FluSight began, the ILI data was the object of the forecasts. The data was released publicly at HHS region levels, and forecast teams were asked to provide forecasts of several ILI targets on the regional levels including season onset, 1-4 week ahead ILI levels, and the week of peak ILI activity \cite[]{biggerstaff2016results,mcgowan2019collaborative}. Currently, the ILI data is collected by the CDC and published on an online portal for viewing at the national, HHS region, census, and state levels \cite[]{cdc2024fluviewportal}. To obtain ILI data, we used the \texttt{R} package \texttt{cdcfluview} which provides functions for downloading the data \cite[]{rudis2021cdcfluview}. Weekly ILI data from the national, HHS region, and census levels are available from the 1997 flu season until the current season. At the state level, data is available from the 2010 flu season to the current season.

Figure \ref{fig:us_ili} shows the ILI data at the national level for flu seasons 2010 to 2023. For most seasons there are 52 weeks listed, but for the 2010, 2015, and 2021 seasons there are 53 weeks because there were 53 Sundays during those seasons. 
To better align with the flu behavior, week 1 is set as the first week of August and week 52 or 53 is the last week in July of the following year. For example week 1 of the 2013 season corresponds to the first week of August 2013, and week 52 of the same season corresponds to the last week of July 2014. This convention is used for the remainder of this manuscript.

\begin{figure}
    \centering
    \includegraphics[scale=.5]{Images/us_ili_seasons.png}
    \caption{Percentage of outpatient visits with an influenza-like illness (ILI) in the US for seasons 2010 to 2023. Week 1 is the first week of August of the year the flu season begins and the last week of the season is the last week of July of the following year.}
    \label{fig:us_ili}
\end{figure}

Notable from the plots in figure \ref{fig:us_ili} is the regular trajectory of the ILI. With the exception of season 2020, the ILI begins low at week 1 and increases as the fall and winter progress until the ILI reaches a peak. As spring progresses to summer, the ILI decreases to low values. As \cite{osthus2019dynamic} point out, there is nearly always either a global or local peak at week 22 which typically corresponds to the week between Christmas and New Year's day. 
Whether local or global, the ILI holiday peak is generally expected and thought to be due to widespread holiday travel, school closure, or other unique social behavior   
\cite[]{ewing2017contact, garza2013effect}.
The only seasons when there was not a peak at week 22 were season 2022 where the season peak occurred particularly early and season 2020 which was greatly influenced by the COVID-19 pandemic.

\subsection{Hospitalization Data}

Hospital admission data, used as the object of FluSight forecasting for the 2022 and 2023 seasons, is based on the CDC's National Healthcare Safety Network (NHSN) dataset entitled \textit{HealthData.gov COVID-19 Reported Patient Impact and Hospital Capacity by State Timeseries}. Several targets of respiratory illnesses including COVID-19, RSV, and influenza are reported weekly by most hospitals in the US. In February 2022 it became mandatory for all hospitals to report the number of COVID-19 and influenza hospitalizations, and since then reporting of hospitalizations has become widespread. These data were updated every Wednesday and Friday according to NHSN guidelines \cite[]{healthdata2024covidts}.

Figure \ref{fig:us_hosp} shows the weekly national hospitalizations for the 2022 and 2023 flu seasons. These plots show similarities to the ILI plots in \ref{fig:us_ili} in that at the early weeks of the season hospitalizations are low, but they increase in the fall to a peak after which they decrease until the flu outbreak ends. For both the 2022 and 2023 seasons, the hospitalizations peaked during the same week as ILI, and in 2023 that peak occurred during the holiday week 22. Additional plots of ILI data and hospitalization
data for select states are included in the supplementary material 
\cite[]{wadsworth2024bas}. These plots show flu outbreak trajectories similar to 
those in figures \ref{fig:us_ili} and \ref{fig:us_hosp} with activity increasing
to a peak and then decreasing.

\begin{figure}
    \centering
    \includegraphics[scale=.5]{Images/us_hospitalizations.png}
    \caption{Weekly flu confirmed hospitalization counts at the national level for 2022 (left) and 2023 (right) flu seasons}
    \label{fig:us_hosp}
\end{figure}

\section{ILI and hospitalization forecast modeling} \label{sec:functions}

% \jarad{Why not start with the ILI model from Osthus et al? It makes a lot more sense to me to start with what has already been done and then expand upon what they have done. Later you have a section called ILI Model, but isn't this the ILI model?}

The typical behavior of the ILI data which starts at lower values in the late summer and fall but which increases to a peak, usually in December or January, followed by a decline motivates the use of a nonlinear function which follows of a similar trajectory for modeling ILI. Compartmental models are standard mathematical models used for disease outbreaks. One important compartmental model is the susceptible-infectious-recovered (SIR) model, which is used by some to model ILI data \cite[]{osthus2019dynamic, allen2017primer}. \cite{ulloa2019} chose to use the asymmetric Gaussian (ASG) function to model ILI data. The SIR and ASG models may both be appropriate to describe ILI behavior over the course of a flu season, however there may also be systematic behavior not captured by either, necessitating an additional model component to capture the discrepancy. In the first part of this section, we present an ILI model similar to the model in \cite{osthus2019dynamic}. With some generalization, the model may incorporate any appropriate nonlinear function, though the focus here is on the SIR and ASG functions which are defined.

With the aim of forecasting hospitalizations, we also introduce a linear model of hospitalizations data with ILI as a predictive covariate. To forecast hospitalizations, ILI data is first forecasted and the forecast is then plugged in as a covariate in the hospitalization model, thus producing hospitalization forecasts. The hospitalization model is also defined in this section, and the section is concluded with descriptions of selected prior distributions, model implementation, and posterior sampling.

\subsection{ILI Model} \label{sec:ili_model}

The proposed model for ILI for any location is given in (\ref{eq:ili_model}). Here $ILI_{s,w}$ is the ILI for flu season $s$ and week $w = 1, 2, ..., W$, where $W = 52$ or $W = 53$, depending on how many Sundays there are in a given season. The ILI is a proportion, so the Beta random variable is a natural selection for modeling. Under the parameterization in of the Beta distribution used in (\ref{eq:ili_model}) the expected value is $\pi_{s,w}$ and the variance is $\pi_{s,w}(1 - \pi_{s,w})/(1 + \kappa_s)$, making $\kappa_s$ a scale parameter. The nonlinear function $f_{\theta_s}(w)$ captures the trajectory of the ILI, and $\gamma_w$ is a discrepancy term for capturing the systematic patterns which $f_{\theta_s}(w)$ does not capture. 
\begin{equation}
\begin{aligned}
    \label{eq:ili_model}
    % \begin{flalign*}
        ILI_{s,w} &\overset{ind}{\sim} \text{Beta}(\pi_{s,w}\kappa_s,\; \kappa_s(1 - \pi_{s,w})) \\
        \text{logit}(\pi_{s,w}) &= f_{\theta_s}(w) + \gamma_w
    % \end{flalign*}
\end{aligned}
\end{equation}
In \cite{osthus2019dynamic} $f_{\theta_s}(w) = \text{logit}(I_{s,w})$ where $I_{s,w}$ is the infectious compartment of the SIR model from (\ref{eq:sir_diff}) in section \ref{sec:sir_func}. In \cite{ulloa2019} $f_{\theta_s}(w) = ASG_{\theta}(w)$ from (\ref{eq:asg_function}) in section \ref{sec:asg_func}. In \cite{ulloa2019} modeling is done hierarchically over seasons.

\subsection{Susceptible-Infectious-Recovered (SIR) compartmental model} \label{sec:sir_func}

The SIR compartmental model is a mathematical model used for modeling disease outbreaks and was introduced by \cite{kermack1927contribution}. Since then, compartmental models have become standard for modeling infectious diseases \cite[]{allen2008mathematical}, and many extensions have been made and studied \cite[for example]{simon2020sir, allen2017primer, van2008deterministic}. The SIR mathematical model includes three compartments and assumes that at any time $t>0$ every individual in a closed population belongs to exactly one compartment. The three compartments are susceptible ($S$), infectious ($I$), and recovered ($R$), and their interaction over the course of an outbreak is described by the differential equations in (\ref{eq:sir_diff}).

\begin{equation}
    \label{eq:sir_diff}
    \frac{dS}{dt} = -\beta SI, \quad \frac{dI}{dt} = \beta S I - \delta I, \quad \frac{dR}{dt} = \delta I
\end{equation}
Here $S$, $I$, and $R$ represent the proportion of the population in each compartment such that $S + I + R = 1$ for all $t$. The trajectory is determined by the disease transmission rate $\beta > 0$ and the recovery rate $\delta > 0$. Respectively, these may be thought of as the expected proportion of susceptible individuals who will be infected by an infectious individual, and the expected rate of recovery to an immune state for a newly infected person. Whether or not a disease outbreak is classified as an epidemic is determined by the initial susceptible population $S_0$, or the susceptible population at time $0$, and the parameter $\rho = \delta/
\beta$. If $S_0/\rho > 1$, the outbreak is considered an epidemic. It is non-epidemic if $S_0/\rho \leq 1$ \cite[]{osthus2019dynamic}. Figure \ref{fig:sir_traj} shows the trajectory of the three compartments of an SIR model where the $S_0$ and $\rho$ were selected to match an outbreak that would classify as epidemic. In the case where $S_0 \leq \rho$, the trajectory for the $I$ compartment would never be increasing. The increase to a peak and subsequent decrease in the $I$ compartment of figure \ref{fig:sir_traj} suggest it is reasonable to model ILI by this compartment. Thus in modeling the ILI data, we consider the data to be analogous to the $I$ proportion of the population.

\begin{figure}
    \centering
    \includegraphics[scale=.6]{Images/sir_traj.png}
    \caption{Susceptible-infectious-recovered (SIR) model separated by compartments. The three compartments are the susceptible compartment (left), infectious (center), and recovered (right). In this example, $S_0/\rho > 1$.}
    \label{fig:sir_traj}
\end{figure}

\subsection{Asymmetric Gaussian (ASG) function} \label{sec:asg_func}
The ASG function is another example of a nonlinear function which can approximate the trajectory of the flu outbreak. The ASG was previously used by Ulloa to model and forecast ILI \cite[]{ulloa2019}, and it has been used to model vegetation growth and satellite sensor data \cite[]{lewis2020extracting, jonsson2002seasonality, hird2009noise, beck2006improved, atkinson2012inter}. The ASG is a modification of the asymmetric Gaussian distribution \cite[]{wallis2014two} and is characterized by its rise to a peak and fall from that peak which may not occur at the same rate, as shown in figure \ref{fig:asg_function}. The ASG function is denoted as $ASG_\theta(w)$ where $\theta = (\lambda, \nu, \mu, \sigma_1^2, \sigma_2^2)$, $\nu > 0$, $\lambda > 0$, $\mu \in (-\infty, \infty)$, $\sigma_1, \sigma_2 > 0$ and $w \in (1, ..., W)$ is week. The function is defined in (\ref{eq:asg_function}).

\begin{figure}
    \centering
    \includegraphics[scale=.45]{Images/asg_function.png}
    \caption{Example plot of asymmetric Gaussian (ASG) function showing the shape of the function in relation to the parameters $\lambda, \; \eta, \; \mu, \; \sigma_1, \; \text{and} \; \sigma_2$}
    \label{fig:asg_function}
\end{figure}

\begin{equation}
    \label{eq:asg_function}
    ASG_{\theta}(w) = 
    \begin{cases}
        \lambda + (\nu - \lambda) \text{exp}[-(w - \mu)^2/2\sigma^2_1], \;\;\; w < \mu \\
        \lambda + (\nu - \lambda) \text{exp}[-(w - \mu)^2/2\sigma^2_2], \;\;\; w \geq \mu
    \end{cases}
\end{equation}
For modeling in this manuscript, we use a slightly reparameterized version of the function in (\ref{eq:asg_function_rep}), where $\eta = \nu - \lambda > 0$. This constraint guarantees that the function has a peak greater than $\lambda$.

\begin{equation}
    \label{eq:asg_function_rep}
    ASG_{\theta}(w) = 
    \begin{cases}
        \lambda + \eta \text{exp}[-(w - \mu)^2/2\sigma^2_1], \;\;\; w < \mu \\
        \lambda + \eta \text{exp}[-(w - \mu)^2/2\sigma^2_2], \;\;\; w \geq \mu
    \end{cases}
\end{equation}

\subsection{Model discrepancy}

The SIR and ASG functions are useful for capturing the main trend of the ILI data, but as \cite{osthus2019dynamic} points out there may be systematic behavior that these or other possible functions may not capture. As noted in section \ref{sec:data}, figure \ref{fig:us_ili} shows a regular peak at week 22 of the flu season. 
Figures \ref{fig:asg_fits} and \ref{fig:discrepancy} are used together to illustrate the systematic discrepancy from a fitted function. Figure \ref{fig:asg_fits} shows the US ILI percentage for all flu seasons from 2010 to 2022 excluding 2020 with a best fit ASG function plotted over the ILI. The fits for each season were made by obtaining the maximum likelihood estimate (MLE) of a model given the ILI data where we assume the ASG function is the mean parameter of a Beta distributed random variable. Figure \ref{fig:discrepancy} shows the discrepancy between the model fit and the data for the same seasons. The grey lines show the difference between the data and the functions from figure \ref{fig:asg_fits} for each season, and the black line is the average by week over all seasons. The lines show that the ASG function typically underpredicts week 22 and overpredicts week 23. Perhaps for other weeks, like week 30 for example, there also tends to be systematic behavior not captured by the ASG function.

\begin{figure}
    \centering
    \includegraphics[scale=.5]{Images/asg_fits.png}
    \caption{Observed US national influenza-like illness (ILI) percentage for seasons 2010 to 2022 excluding 2020 (grey) overlaid with MLE of an asymmetric Gaussian (ASG) model for the ILI data (black)}
    \label{fig:asg_fits}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[scale=.45]{Images/discrepancy.png}
    \caption{Difference between observed US national influenza-like illness (ILI) and MLE fits for an asymmetric Gaussian (ASG) model for each season 2010 to 2022 excluding 2020 (grey) and the average difference of all seasons (black)}
    \label{fig:discrepancy}
\end{figure}

The term $\gamma_w$, where $w$ is the season week, is included in (\ref{eq:ili_model}) to capture the per week discrepancy between ILI and the function. Modeling discrepancy has been used in uncertainty analysis of simulators to capture systematic differences between mathematical models and 
reality \cite[]{ma2022multifidelity,brynjarsdottir2014learning,arendt2012improving,kennedy2001bayesian}. Modeling discrepancy can lead to overfitting, particularly in forecasting scenarios, and may also lead to identifiability issues. 
Thus, care must be taken in setting parameter constraints as well as in the selection of prior distributions \cite[]{osthus2019dynamic,brynjarsdottir2014learning}. 
Modeling of the discrepancy for ILI was done by \cite{osthus2019dynamic} during the 2015 and 2016 flu seasons where their model outperformed all others in the CDC flu forecasting challenge \cite[]{osthus2019dynamic}. As in Osthus's model, $\gamma_w$ is modeled as a reverse random walk, as shown in (\ref{eq:rw}). 
\begin{equation}
    \label{eq:rw}
        \gamma_w|\gamma_{w + 1} \overset{ind}{\sim} N(\gamma_{t+1},\sigma^2_{\gamma}), \quad
        \gamma_{W} \sim N(0,\sigma^2_{\gamma_W})
\end{equation}
The idea for using the reverse random walk is that there are several previous seasons of ILI data, and assuming the random walk captures systematic behavior, fitting it hierarchically over seasons can assist in predicting future behavior in the current season. 
Reverse random walks have also been used with success in election forecasting and other flu forecasting models \cite[]{osthus2021multiscale, osthus2019dynamic, linzer2013dynamic}. The sum to zero constraint $-\gamma_1 = \sum_{w=2}^W \gamma_w$ is imposed on (\ref{eq:rw}) to improve identifiability.

\subsection{ILI forecasts}
Model (\ref{eq:ili_model}) is fit via Bayesian posterior updating. Future ILI forecasts are obtained via the posterior predictive distribution where for week $w$, the predictive distribution is obtained by integrating over the parameters $\pi_s$, $\kappa_s$, $\sigma^2_{\gamma}$, and $\sigma^2_{\gamma_W}$ as in (\ref{eq:ili_post}) where $p(\boldsymbol{\pi}_s, \kappa_s, \sigma^2_{\gamma}, \sigma^2_{\gamma_W} | \textbf{ILI})$ is the density function of the posterior distribution for the model parameters. If the current week is $w^*$ then the desired forecasts are for weeks $w^* + i$ where $i$ is a positive integer.

\begin{equation}
    \label{eq:ili_post}
    p(\widetilde{ILI}_{s,w^*} | \textbf{ILI}) = \int \int \int \int p(\widetilde{ILI}_{s,w^*} | \boldsymbol{\pi}_s, \kappa_s, \sigma^2_{\gamma}, \sigma^2_{\gamma_W}) p(\boldsymbol{\pi}_s, \kappa_s, \sigma^2_{\gamma}, \sigma^2_{\gamma_W} | \textbf{ILI}) d\boldsymbol{\pi}_s d \kappa_s d \sigma^2_{\gamma} d \sigma^2_{\gamma_W}
\end{equation}
% \jarad{It doesn't make sense to have 4 $\int$ here since $\pi_s$ is a vector.}
% \spencer{Isn't that captured in $d \pi_s$?}

\subsection{Hospitalization model} \label{sec:hospital_model}

The second component for forecast modeling is based on the relationship between hospitalizations and ILI and is defined in (\ref{eq:final_hosp}).
This is an example of an autoregressive model with exogenous variables where the autoregressive lag is one (ARX(1)) \cite[]{raftery2010online,ljung1987system}. Figure \ref{fig:lag_scatter}
shows scatterplots of the difference between hospitalizations and scaled 1 week 
lags by ILI percentage at the US national level. The points in the left plot are
from hospitalization differences whereas in the right the hospitalizations were
log transformed. The positive and slightly curved relationship suggests model
(\ref{eq:final_hosp}) is a reasonable choice.

\begin{figure}
    \includegraphics[scale=.46]{Images/lagdiff_vs_ili.png}
    \caption{Scatterplots of national hospitalizations (left) or log hospitalizations (right) lag differences by ILI \% where the lag is scaled by $\phi$. Points are colored by season. For hospitalizations $\phi = 0.6$, and for log hospitalizations $\phi = 0.77$. }
    \label{fig:lag_scatter}
\end{figure}

\begin{equation}
    \begin{aligned}
    \label{eq:final_hosp}
    H_{s,w} &= \alpha_{0s} + \alpha_{1s} (ILI_{s,w} \times P) + \alpha_{2s} (ILI_{s,w} \times P)^2 + \phi H_{s,w-1} + \epsilon_{s,w}\\ 
    \epsilon_{s,w} &\overset{iid}{\sim} D_s(0, \sigma_{\epsilon_s}^2 \times P, \omega_s) %N(0, \sigma_{\epsilon}^2)
    \end{aligned}
\end{equation}
In model (\ref{eq:final_hosp}), $H_{s,w}$ is the number of hospitalizations for week $w$ in season $s$, $\epsilon_{s,w}$ is an error term distributed according to some distribution $D_s$ with mean parameter 0, scale parameter $\sigma_{\epsilon_s}$, and the additional parameter $\omega_s$ is the degrees of freedom parameter when $D_s$ belongs to the location-scale t (LST) family.

This model is for any location, and for fitting purposes $ILI_{s,w}$ is always multiplied by $P$ which is proportional to the population of the state or territory, in this case the total population divided by 50,000. This is done as a means of scaling so that the prior distribution assigned to  $\boldsymbol{\alpha}_s = (\alpha_{0s}, \alpha_{1s}, \alpha_{2s})$, $\sigma_{\epsilon_s}$, and $\omega_s$ might reasonably be the same for all states.

Like the ILI model, the hospitalization model in (\ref{eq:final_hosp}) is also fit via Bayesian posterior updating. To obtain forecasts for $H_{s, w^* + i}$, the ILI posterior predictive distribution is used along with the posterior distribution for the parameters in (\ref{eq:final_hosp}). We considered three scenarios for model (\ref{eq:final_hosp}). A model where $D_s$ belongs to the normal family (NORM), $D_s$ belongs to the LST family, and one where $H_{s,w}$ is replaced with $\text{log}(H_{s,w} + 1)$ and $D_s$ is from a normal family, or $H_{s,w} + 1$ is lognormally distributed (LNORM). The population value $P$ is excluded from the LNORM model, and we set $\boldsymbol{\alpha}_{22} = \boldsymbol{\alpha}_{23}$ to help with fitting. In the LNORM model, if the linear parameters are not set the same for seasons 22 and 23, the final variance was more prone to be extreme. Besides varying the distribution family of hospitalizations, we also considered $\alpha_{2s} = 0$ or there is no quadratic ILI term.

\subsection{Prior selection}

The priors selected for the ILI data model under both the SIR and ASG models largely follow the prior selections in \cite{osthus2019dynamic} and \cite{ulloa2019} with a few exceptions where changes improved numerical stability and/or we felt the adjusted prior made more sense for the problem. For model (\ref{eq:ili_model}), parameters which are common even when using different functions of $f_{\theta_s}(w)$ are $\kappa_s$, $\sigma_{\gamma}^2$, and $\sigma_{\gamma_W}^2$. For the SIR function $\theta_s = (S_{0s}, I_{0s}, R_{0s}, \alpha_s, \rho_s)$, and for the ASG function $\theta_s = (\alpha_s, \eta_s, \mu_s, \sigma_{1s}^2, \sigma_{2s}^2)$. For the hospitalization model in (\ref{eq:final_hosp}) the parameter to be estimated is $\Psi = (\alpha_{0s}, \alpha_{1s}, \alpha_{2s}, \phi, \sigma_{\epsilon_s}, \omega_s)$.

The priors assigned were mostly noninformative, though in certain cases the prior distributions were selected for numerical stability as was the case for $\sigma_{\gamma}^2$ and $\sigma_{\gamma_W}^2$. For these two scale parameters only, rather than assigning a half-normal prior to the standard deviation parameters, as recommended by \cite{gelman2006prior}, the priors were assigned to the variance parameters. Univariate parameters were assigned either a normal distribution prior if the support is on $\mathbb{R}$, a half-normal prior if the support is nonnegative, or a truncated-normal prior to match a more specific support. Under the ASG model, $\theta_s$ is modeled hierarchically over seasons so that for each season the transformed parameter $T(\theta_s) \sim N(\theta, \Sigma$) and priors distributions are assigned to $\theta$ and $\Sigma$.

Additional prior constraints were made to improve parameter identifiability. In \cite{osthus2019dynamic} the initial value of the susceptible population compartment of the SIR model was set to $S_0 = 0.9$. The parameters $I_{0s}$, $\beta_s$, and $\rho_s$ were assigned informative priors. To improve identifiability when $f_{\theta_s}(w) = ASG_{\theta_s}(w)$ in (\ref{eq:ili_model}) we followed a modular Bayesian approach for fitting the parameters. A modular Bayesian approach involves multiple steps of parameter fitting where some parameters may be estimated without priors via maximum likelihood estimation or other means. Fitting the rest of the model parameters involves assigning priors and conditioning on the previously fit parameters. This has been done in computer modeling to improve identifiability and other issues, though \cite{liu2009modularization} warn this approach is not probabilistically sound if parameter inference is a priority \cite[]{jiang2015surrogate, arendt2012improving, liu2009modularization}. We carried out the modular fit by first estimating the maximum likelihood estimate (MLE) for the parameter $\lambda_s$ for each season in (\ref{eq:asg_function_rep}) and plugging in the MLE as a fixed value.

% \spencer{Blue stuff below will be either included further on or will be removed entirely

\subsection{Parameter estimation and posterior predictive sampling}
\label{sec:implementation_posterior}

>>>>

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Only one address is permitted per author. %%
%% Only division, organization and e-mail is %%
%% included in the address.                  %%
%% Additional information can be included in %%
%% the Acknowledgments section if necessary. %%
%% ORCID can be inserted by command:         %%
%% \orcid{0000-0000-0000-0000}               %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\snm{Spencer Wadsworth}\ead[label=e1]{sgw96@iastate.edu}}
% \author[A]{\fnms{Second}~\snm{Author}\ead[label=e2]{second@somewhere.com}\orcid{0000-0000-0000-0000}}
\and
\author[A]{~\snm{Jarad Niemi}\ead[label=e2]{niemi@iastate.edu}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Department of Statistics,
Iowa State University, Ames, IA 50011}

@@

\input{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% `\ ' is used here because TeX ignores    %%
%% spaces after text commands.              %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\input{manuscript_res}
